---
title: "Intro to R Bootcamp 2026: Start Here"
author: "UM Biostatistics"
format:
  html:
    toc: true
    toc-depth: 2
execute:
  echo: true
  warning: false
  message: false
---

## Goal

By the end of this project, you will have a small reproducible analysis that imports data, checks it, cleans it, summarizes it, plots it, and fits a simple model.

This dataset is simulated for teaching. It does not contain real people.

## 1. Setup

```{r}
# Run scripts/00_setup.R first if tidyverse or broom is not installed.
library(tidyverse)
library(broom)
```

AI checkpoint: Ask an AI assistant to explain what `library(tidyverse)` does. Then verify by running `search()` or `?library` in R.

## 2. Import

```{r}
health_raw <- read_csv("../data/bootcamp_health.csv")
```

```{r}
glimpse(health_raw)
```

## 3. Inspect

Use small checks before changing the data.

```{r}
# How many rows and columns?

# What are the first few rows?

# How many missing values are in each column?

```

## 4. Clean and create variables

Create a cleaned dataset called `health_clean`.

```{r}
health_clean <- health_raw |>
  mutate(
    visit_date = as.Date(visit_date),
    hypertension_flag = systolic_bp >= 130 | diastolic_bp >= 80,
    bmi_group = case_when(
      is.na(bmi) ~ NA_character_,
      bmi < 18.5 ~ "Underweight",
      bmi < 25 ~ "Healthy range",
      bmi < 30 ~ "Overweight",
      TRUE ~ "Obesity range"
    )
  ) |>
  filter(!is.na(age), !is.na(bmi), !is.na(systolic_bp))
```

AI checkpoint: Ask AI to explain one line of the `case_when()` code. Do not ask it to do the whole lab.

## 5. Summarize

```{r}
health_clean |>
  group_by(sex) |>
  summarize(
    n = n(),
    mean_age = mean(age),
    mean_bmi = mean(bmi),
    percent_hypertension = mean(hypertension_flag) * 100,
    .groups = "drop"
  )
```

## 6. Plot

Build the plot one layer at a time.

```{r}
health_clean |>
  ggplot(aes(x = bmi, y = systolic_bp)) +
  geom_point()
```

Now add color, a trend line, labels, and a theme.

```{r}
# Your improved plot here
```

## 7. Simple model

The goal is to learn syntax and output-reading, not regression theory.

```{r}
fit <- lm(systolic_bp ~ age + bmi + sex, data = health_clean)
summary(fit)
```

```{r}
tidy(fit)
```

AI checkpoint: Ask AI to explain the output in beginner language. Then verify the variable names and sample size yourself.

## 8. Save one output

```{r}
cohort_table <- health_clean |>
  group_by(sex) |>
  summarize(
    n = n(),
    mean_bmi = mean(bmi),
    mean_sbp = mean(systolic_bp),
    .groups = "drop"
  )

write_csv(cohort_table, "output/cohort_table.csv")
```

## Exit check

Write one sentence answering each question:

1. What file did you import?
2. What did you check before cleaning?
3. What did `mutate()` create?
4. What did `group_by()` change about the summary?
5. What would you ask AI to explain, and how would you verify the answer?
