---
title: "Day 2 Workshop: Debugging and Functions"
author: "Mike Kleinsasser and Jacob Gladfelter"
format:
  html:
    toc: true
    toc-depth: 2
execute:
  echo: true
  warning: false
  message: false
  error: true
---

## Goal for this lab

Today is about getting unstuck without starting over.

By the end, you should be able to:

1. Use a repeatable debugging workflow.
2. Translate common R errors into object checks.
3. Turn repeated analysis code into a small function.
4. Complete a short analysis that cleans data, plots, and reads a simple model call.
5. Use documentation, search, and AI responsibly.

This dataset is simulated for teaching. It does not contain real people.

## 0. Setup

Run this chunk first. If it fails, read the error and use the debugging ladder before changing anything else.

```{r}
library(tidyverse)
library(broom)
```

Import the teaching dataset.

```{r}
health <- read_csv("data/bootcamp_health.csv")
glimpse(health)
```

Quick orientation:

```{r}
nrow(health)
names(health)
```

## 1. The debugging ladder

When code breaks, do not rewrite the whole script. Move through the ladder.

1. **Reproduce**: Can I make the error happen again?
2. **Read**: What does the message literally say?
3. **Isolate**: Which line first fails?
4. **Inspect**: What are the object names, types, dimensions, and missing values?
5. **Test**: Which assumption was false?
6. **Verify**: Did the fix solve the original problem?

Use this small log for each bug below.

| Step | Your notes |
|---|---|
| Error message |  |
| First failing line |  |
| Object or column checked |  |
| Smallest fix tried |  |
| Verification result |  |

## 2. Broken-code lab

Run one broken example at a time. For each one:

- Predict what might go wrong.
- Run only the broken code.
- Read the error or surprising result.
- Inspect one object.
- Write the smallest fix in the blank chunk.

### Bug 1: path problem

```{r}
#| eval: false
health_path_bug <- read_csv("bootcamp_health.csv")
```

What should you inspect first?

```{r}
# Try: getwd(), list.files(), list.files("data")
```

Write the fixed code here.

```{r}
# health_path_bug <-
```


### Bug 2: missing package or function

This one is easiest to see in a fresh R session before loading packages.

```{r}
#| eval: false
health_package_bug <- fread("data/bootcamp_health.csv")
```

If you see `could not find function`, what package probably provides the function?

```{r}
# Try: ?read_csv after loading tidyverse or readr
# Try: library(data.table)
```

Write the fixed setup plus import code here.

```{r}
# library(...)
# health_package_bug <-
```

### Bug 3: missing object

```{r}
#| eval: false
health_clean |>
  count(cohort)
```

What object names exist right now?

```{r}
# Try: ls()
```

Write the fixed code here.

```{r}
# Create health_clean or use the object that already exists.
```

### Bug 4: syntax problem

```{r}
#| eval: false
health |>
  summarize(
    mean_bmi = mean(bmi, na.rm = TRUE)
    mean_sbp = mean(systolic_bp, na.rm = TRUE)
  )
```

What does R usually mean when it says something like `unexpected end of input` or waits with a `+` prompt?

Write the fixed code here.

```{r}
# health |>
#   summarize(
#     ...
#   )
```

### Bug 5: incorrect type

```{r}
#| eval: false
health |>
  mutate(days_since_visit = "2026-08-27" - visit_date) |>
  select(participant_id, visit_date, days_since_visit)
```

Inspect the type of `visit_date`.

```{r}
# Try: class(health$visit_date)
```

Write the fixed code here.

```{r}
# health |>
#   mutate(visit_date = as.Date(visit_date), ...)
```

### Bug 6: grouping mistake

This code runs, but the answer is not the intended overall mean.

```{r}
#| eval: false
health_by_cohort <- health |>
  group_by(cohort)

health_by_cohort |>
  summarize(overall_mean_bmi = mean(bmi, na.rm = TRUE))
```

What is still attached to `health_by_cohort`?

```{r}
# Try: groups(health_by_cohort)
```

Write code for the overall mean and, separately, the cohort-specific means.

```{r}
# Overall mean:

# Cohort-specific means:
```

### Bug 7: plotting error

```{r}
#| eval: false
health |>
  ggplot(aes(x = bmi, y = systolic_bp, color = gender)) +
  geom_point()
```

Which variable name is wrong? How can you check?

```{r}
# Try: names(health)
```

Write the fixed plot here.

```{r}
# health |>
#   ggplot(...) +
#   geom_point()
```

### Bug 8: missing values hide a result

This code runs, but the result is not useful.

```{r}
#| eval: false
health |>
  summarize(mean_sbp = mean(systolic_bp))
```

How many missing systolic blood pressure values are there?

```{r}
# health |>
#   summarize(missing_sbp = sum(is.na(systolic_bp)))
```

Write the fixed summary here.

```{r}
# health |>
#   summarize(mean_sbp = mean(...))
```

### Bug 9: model formula error

```{r}
#| eval: false
health |>
  filter(!is.na(bmi), !is.na(systolic_bp)) |>
  lm(systolic_bp ~ age + bmi + smoking, data = _)
```

There are two things to discuss here:

1. What variable name is not in the data?
2. Is this pipe style easy for a beginner to debug?

Write the clearer fixed version here.

```{r}
# health_model_data <- health |>
#   filter(...)
#
# fit_sbp <- lm(..., data = health_model_data)
# summary(fit_sbp)
```

## 3. From repeated code to a function

Start with code that already works.

```{r}
health |>
  group_by(cohort) |>
  summarize(
    n = n(),
    mean_bmi = mean(bmi, na.rm = TRUE),
    sd_bmi = sd(bmi, na.rm = TRUE),
    missing_bmi = sum(is.na(bmi)),
    .groups = "drop"
  )

health |>
  group_by(cohort) |>
  summarize(
    n = n(),
    mean_sbp = mean(systolic_bp, na.rm = TRUE),
    sd_sbp = sd(systolic_bp, na.rm = TRUE),
    missing_sbp = sum(is.na(systolic_bp)),
    .groups = "drop"
  )
```

Before writing a function, answer:

- What stayed the same?
- What changed?
- What should the function return?

### Function scaffold

Complete the function. Keep it small and readable.

```{r}
summarize_numeric_by_group <- function(data, group_col, value_col, na_rm = TRUE) {
  data |>
    group_by({{ group_col }}) |>
    summarize(
      n = n(),
      mean = mean({{ value_col }}, na.rm = na_rm),
      sd = sd({{ value_col }}, na.rm = na_rm),
      missing = sum(is.na({{ value_col }})),
      .groups = "drop"
    )
}
```

Test one simple call first.

```{r}
summarize_numeric_by_group(health, cohort, bmi)
```

Now test a second call.

```{r}
# summarize_numeric_by_group(health, smoking_status, systolic_bp)
```

Add a default argument or a small improvement. Choose one:

- Add `digits = 1` and round the mean and SD.
- Add `min_n = 1` and filter out groups with fewer than `min_n` rows.
- Rename the output columns after the value being summarized.

```{r}
# Improved function here
```

## 4. End-to-end mini-task

Your task is to build a short analysis from scratch. Use small steps and verify each one.

### Step A: clean

Create `health_day2` with:

- `visit_date` converted to a Date.
- `high_sbp`, where systolic blood pressure is at least 130.
- `diabetes_yes`, where diabetes is equal to "Yes".
- `sleep_check`, where sleep is missing, below 6 hours, between 6 and 10 hours, or above 10 hours.
- Complete cases for `age`, `bmi`, `systolic_bp`, and `smoking_status`.

```{r}
health_day2 <- health |>
  mutate(
    # your variables here
  ) |>
  filter(
    # your complete-case filters here
  )

# Verify:
# glimpse(health_day2)
# count(health_day2, sleep_check)
```

### Step B: summarize

Use your function to summarize BMI by smoking status.

```{r}
# summarize_numeric_by_group(...)
```

### Step C: plot

Make a plot of BMI and systolic blood pressure. Add one feature that makes the plot easier to read.

```{r}
# health_day2 |>
#   ggplot(...) +
#   geom_point(...) +
#   ...
```

### Step D: read a model call

Fit a simple linear model. The goal is to read the call and output, not to make causal claims.

```{r}
fit_sbp <- lm(
  systolic_bp ~ age + bmi + smoking_status,
  data = health_day2
)

summary(fit_sbp)
tidy(fit_sbp)
```

In one sentence each:

1. What is the outcome?
2. What are the predictors?
3. Which dataset did the model use?
4. How many rows were used? How did you check?

### Optional GLM stretch

Create a logistic regression for `diabetes_yes`. First make sure the outcome is logical or 0/1.

```{r}
fit_diabetes <- glm(
  diabetes_yes ~ age + bmi + smoking_status,
  data = health_day2,
  family = binomial()
)

tidy(fit_diabetes)
```

What changes in the model call when you move from `lm()` to `glm()`?

## 5. Documentation, search, and AI responsibly

Use AI and search as debugging tools, not as substitutes for running checks.

A good debugging prompt:

> I am a beginning R user. This code gives the error below. Please explain what the error means, name the first object or column I should inspect, and suggest the smallest fix. Do not rewrite the full analysis. Code: ... Error: ...

Before asking for help, prepare a minimal example:

```{r}
# Not the full data; just enough structure to show the problem.
example_health <- tibble(
  cohort = c("screening", "follow-up"),
  bmi = c(24.1, NA),
  systolic_bp = c(118, 135)
)
```

Responsible-use checklist:

- Do not paste protected health information, credentials, or sensitive unpublished data into third-party AI tools.
- Share the error, the relevant line, and fake or simulated example data.
- Verify suggestions by running code and checking objects with `str()`, `glimpse()`, `names()`, and `?function_name`.
- Ask for help when you cannot isolate the failing line, when a result changes unexpectedly, or when the code affects real analysis decisions.

## 6. Competency check

Before leaving, make sure you can do each item with the project dataset.

- [ ] Import `data/bootcamp_health.csv` from the project root.
- [ ] Use the debugging ladder on one broken code example.
- [ ] Fix one object-name or column-name error by inspecting `ls()` or `names()`.
- [ ] Fix one type error by inspecting `class()` or `glimpse()`.
- [ ] Explain why grouped data can produce a surprising summary.
- [ ] Write and test a small summary function.
- [ ] Make one plot and fit one simple `lm()` call.
- [ ] Explain what you asked AI or search, and how you verified the answer.

## Follow-up resources in this project

- `references/debugging_ladder.md`
- `references/dplyr_ggplot_reference.md`
